---
title: "LightGBM based on Sillen script"
author: "Matthew Brousil"
date: "2025-12-08"
output: html_document
---

Code is based on Sillen's code [here](https://github.com/GardnerLab-Pitt/OhioRiver_SR/blob/9f4d8d5d8f82b89b999081d22c944977043dc520/04_LC02_model_train_eval.Rmd) with gpt-4o assistance

```{r load-packages, message = FALSE, warning=FALSE}
library(lightgbm)
library(tidyverse)
library(Metrics)
library(arrow)
library(sf)
library(CAST)
library(tictoc)
```

## Preparation

Read in the Colorado Basin matchups
```{r load-data}
# Gardner corrections
co_matchups_gardner <- arrow::read_feather(
  file = "out/co_basin_matchups_gardner_correction.feather"
)
```

Function definition from Sam Sillen's code
```{r def-funs}
# https://github.com/GardnerLab-Pitt/OhioRiver_SR/blob/9f4d8d5d8f82b89b999081d22c944977043dc520/04_LC02_model_train_eval.Rmd

# This function randomly samples match-ups across different locations, times, and concentrations for splitting training/validation data

holdout <- function(x) {
  
  x <- x %>%
    # Split up into spatial and temporal groups
    group_by(long_group, time_group) %>% 
    dplyr::mutate(mag = cut(harmonized_value, quantile(
      x = harmonized_value,
      c(0, 0.2, 0.4, 0.6, 0.8, 0.9, 1),
      include.lowest = T
    )),
    mag = factor(
      mag,
      labels = c( 0.2, 0.4, 0.6, 0.8, 0.9, 1)
    )) %>%
    ungroup()
  
  set.seed(22)
  
  train <- x %>%
    group_by(time_group, long_group, mag) %>%
    # 80% of data will be used for training
    sample_frac(.8) %>% 
    ungroup() %>%
    dplyr::mutate(.partitions = 1)
  
  validate <- x %>%
    anti_join(train) %>%
    dplyr::mutate(.partitions = 2)
  
  out <- train %>%
    bind_rows(validate) 
  
  return(out)
}
```

<br>

#### Data prep

Filters:

+ harmonized_value > 0.01
+ harmonized_discrete_depth_value <= 10 | is NA

```{r filter-data}
model_prep_data <- co_matchups_gardner %>%
  filter(
    harmonized_value > 0.01,
    # depth < 10m, allow NAs because there are a lot
    harmonized_discrete_depth_value <= 10 |
      is.na(harmonized_discrete_depth_value)
  ) %>% 
  select(siteSR_id, date, lat, lon, huc8, tier, harmonized_value,
         prop_hillShadow, mean_hillShade, Red, Green, Blue, Nir, Swir1, Swir2,
         Surface_temp_kelvin, NR:saturation, dw) %>%
  mutate(
    year = year(date),
    log_harmonized_value = log(harmonized_value))
```

<br>

Temporal distribution
```{r plot-years}
model_prep_data %>%
  ggplot() +
  geom_histogram(aes(year), color = "black", fill = "white", binwidth = 1) +
  theme_bw()
```

## Model building

### Train/test holdout
```{r create-holdouts}
# https://github.com/GardnerLab-Pitt/OhioRiver_SR/blob/9f4d8d5d8f82b89b999081d22c944977043dc520/04_LC02_model_train_eval.Rmd
model_pre_train_prep <- model_prep_data %>%
  mutate(
    # Sam Sillen code:
    lat_group = cut_number(lat, 5, right= F),
    long_group = cut_number(lon, 5, right=F),
    # date = lubridate::ymd(date),
    julian = as.numeric(julian.Date(date)),
    space_group = paste0(lat_group, long_group),
    time_group = cut_number(julian, 5, right=F)
  ) %>%
  holdout() %>% 
  ungroup() %>%
  as.data.frame() %>%
  filter_all(all_vars(!is.infinite(.))) %>%
  filter_all(all_vars(!is.nan(.)))

# Training data, 80%
train <- model_pre_train_prep %>%
  filter(.partitions == 1)

# Test set
holdout <- model_pre_train_prep %>%
  filter(.partitions == 2)

# Vars to use
# Features that were in top 20 of previous model run var importance
features <- c("B.GN", "G.BR", "GR2", "IRG", "KRL", "R.BS", "Surface_temp_kelvin",
              "B.RG", "BG", "SR", "EVI", "GB", "GCI", "GN2", "N_S", "SN", "B.GS",
              "B.NS", "G.BN", "G.BS")

# Create cross validation folds for spatial-temporal cross validation
folds <- CreateSpacetimeFolds(train, spacevar = "long_group",
                              timevar = "time_group", k = 5)
```

### Training
```{r train_model}
# Separate features and target variable
train_x <- train %>%
  select(all_of(features)) %>%
  as.matrix()

train_y <- train$log_harmonized_value

# Create LightGBM datasets
dtrain <- lgb.Dataset(data = train_x, label = train_y)

# Step 4: Define model parameters
params <- list(
  objective = "regression",  # Continuous response variable
  metric = "rmse",            # Root Mean squared error (RMSE)
  boosting = "gbdt",        # Gradient Boosting Decision Tree
  learning_rate = 0.1,      # Learning rate
  lambda = 0
)

param_grid <- expand.grid(
  # eta
  learning_rate = c(0.05, 0.1, 0.3),
  num_leaves = c(15, 31),
  max_depth = c(-1, 5),
  alpha = c(0.01, 0.1, 0.5, 1),
  lambda = c(0.01, 0.1, 0.5, 1)
)

# Loop through each combination of hyperparameters
cv_results <- list()


tic()

set.seed(22)

for (i in 1:nrow(param_grid)) {
  params_temp <- list(
    objective = "regression",
    metric = "rmse", 
    learning_rate = param_grid$learning_rate[i],
    num_leaves = param_grid$num_leaves[i],
    max_depth = param_grid$max_depth[i],
    alpha = param_grid$alpha[i],
    lambda = param_grid$lambda[i],
    seed = 122
  )
  
  # Perform cross-validation
  cv_model <- lgb.cv(
    params = params_temp,
    data = dtrain,
    folds = folds$indexOut,
    nrounds = 500,                  # Number of boosting iterations
    early_stopping_rounds = 10,     # Stop early if no improvement
    verbose = -1                   # Print training progress
  )
  # Store results
  cv_results[[i]] <- list(
    params = params_temp,
    best_iter = cv_model$best_iter,
    best_rmse = min(unlist(cv_model$record_evals$valid$rmse$eval))
  )
}

toc()

# Find the best combination of hyperparameters
best_result <- cv_results[[which.min(sapply(cv_results, function(x) x$best_rmse))]]
print(best_result)

# Train the final model using the best hyperparameters and number of iterations
final_model <- lgb.train(
  params = best_result$params,
  data = dtrain,
  nrounds = best_result$best_iter,
  verbose = 1
)

lgb.plot.importance(lgb.importance(final_model))
```

### Testing
```{r predict-holdout-set}
holdout_x <- holdout %>%
  select(all_of(features)) %>%
  as.matrix()

holdout_y <- holdout$log_harmonized_value

# Predict from candidate models:
predictions <- predict(final_model, holdout_x)

# Combine the held out data and the predictions
evaluation_data <- data.frame(
  actual = holdout_y,
  predicted = predictions,
  date = holdout$date,
  siteSR_id = holdout$siteSR_id
)
```

### Holdout prediction
```{r plot-holdout-performance}
evaluation_data %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Holdout: Predicted vs Actual Values",
       x = "Actual Values", y = "Predicted Values") +
  xlim(c(0, max(c(evaluation_data$actual, evaluation_data$predicted)))) +
  ylim(c(0, max(c(evaluation_data$actual, evaluation_data$predicted)))) +
  theme_bw()
```


### Full predicted vs. actual 1:1
```{r predicted-vs-actual}
training_predictions <- predict(final_model, train_x)

train_predicted <- data.frame(
  actual = train_y,
  predicted = training_predictions,
  date = train$date,
  siteSR_id = train$siteSR_id
)

# Tag predictions with whether they were used to train or not
full_predicted <- bind_rows(
  train_predicted %>%
    mutate(origin = "Training set"),
  evaluation_data %>%
    mutate(origin = "Holdout set")
)

full_predicted %>%
  ggplot() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  geom_point(aes(x = actual, y = predicted, fill = origin),
             pch = 21, color = "black") +
  labs(title = "Full dataset: Predicted vs. Actual values",
       x = "Actual Values", y = "Predicted Values") +
  xlim(c(0, max(c(full_predicted$actual, full_predicted$predicted)))) +
  ylim(c(0, max(c(full_predicted$actual, full_predicted$predicted)))) +
  theme_bw()
```

<br>

### Timeseries plotting

#### Training timeseries
```{r training-time-series-plot, fig.height = 9, fig.width=7}
# Plot of a subset of the biggest sites
biggest_sites <- full_predicted %>%
  count(siteSR_id) %>%
  arrange(desc(n)) %>%
  head(n = 10) %>%
  pull(siteSR_id)

full_predicted %>%
  filter(
    siteSR_id %in% biggest_sites,
    origin == "Training set"
  ) %>%
  ggplot(aes(x = date)) +
  # Observed
  geom_line(aes(y = actual, linetype = "Observed"), color = "orange") +
  # Predicted
  geom_point(aes(y = predicted, size = "Predicted"),
             pch = 21, color = "black", fill = "salmon") +
  facet_wrap(vars(siteSR_id), scales = "free", ncol = 2) +
  ggtitle("Predicted & actual values of top 10 most populous sites",
          subtitle = "Training set only") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

#### Holdout timeseries
```{r holdout-time-series-plot, fig.height = 9, fig.width=7}
# Plot of a subset of the biggest sites
biggest_sites <- full_predicted %>%
  count(siteSR_id) %>%
  arrange(desc(n)) %>%
  head(n = 10) %>%
  pull(siteSR_id)

full_predicted %>%
  filter(
    siteSR_id %in% biggest_sites,
    origin == "Holdout set"
  ) %>%
  ggplot(aes(x = date)) +
  # Observed
  geom_line(aes(y = actual, linetype = "Observed"), color = "orange") +
  # Predicted
  geom_point(aes(y = predicted, size = "Predicted"),
             pch = 21, color = "black", fill = "salmon") +
  facet_wrap(vars(siteSR_id), scales = "free", ncol = 2) +
  ggtitle("Predicted & actual values of top 10 most populous sites",
          subtitle = "Holdout set only") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

### Residuals
```{r residual-plot}
full_predicted %>%
  mutate(residuals = actual - predicted) %>%
  ggplot(aes(x = predicted, y = residuals)) +
  # geom_point(alpha = 0.6) +
  geom_hex() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(vars(origin)) +
  labs(title = "Residuals vs Predicted Values", x = "Predicted", y = "Residuals") +
  scale_fill_viridis_c() +
  theme_bw()
```

